\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed

### 2015-06-18 Paper_JFM/grass.tex
Add discussion on Mode 2

Next we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in ~\eqref{navier-stokes} and expand to linear order to investigate the evolution of small perturbations $(\tilde{u}, \tilde{v})$, which obey
\begin{equation}
\begin{split}
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0,
\end{split} \nonumber
\end{equation}
where the tilde are dropped.
These equations are non-dimensionalized using half channel height $H$, velocity $U_0$, and time $H/U_0$, leading to three non-dimensional parameters, \textit{viz.} $\Rey$, $\Ndg$, and the vegetation submergence ratio $\hg/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the vegetation density and help elucidate the instability mechanism. 
Using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a solution of 
the form $\left(u,v,\psi \right)= \left(\hat u(y), \hat v(y), \phi(y) \right)e^{ikx+\sigma t}$ to obtain a modified Orr-Sommerfield equation \cite{Drazin81} 
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+D\left(2R \Ndg S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$. 
The growth rate $\sigma$ for a given wave number $k$ appears as an eigenvalue that allows a non-trivial solution $\phi$ of  \eqref{Orr-somerfield}.
We solve \eqref{Orr-somerfield} numerically for $\sigma$ and $\phi$.

@@

\begin{table*}
% \footnotesize
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& yes, near $y=\hg$			& no, spans water column
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two unstable modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the instability mechanism, we consider the dependence of the fastest growing wavenumber on $\delta$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical $\delta$ discontinuously jumps and remains $O(1)$ (see Fig.~\ref{Re_vs_delta}). 
To aid in explaining this behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in Fig.~\ref{K_Re_sigma_set3}. 
The smallest $\Rey$ on the neutral curve (Re$(\sigma)=0$) sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two; we refer to the region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

@@

\subsection{Mode 2}
Similarly, the threshold condition for Mode 2 is numerically observed to be $\Rey \propto ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in Fig.~\ref{Re_vs_delta}, which can be understood by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
% Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the vegetation is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
In this limit ~\eqref{Orr-somerfield} simplifies to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for $y>\hg$}.
\end{split}
\label{eqn:mode2asymp}
\end{align}
\begin{figure}
\includegraphics[]{Asymptotic_noshear}
% \includegraphics[width=12cm]{Asymptotic_noshear}
\caption{
Plot of the neutral Mode 1 (solid) and Mode 2 (dashed) shape $|\phi|$ in the limit of small $\delta/H$ for $\hg/H=0.2$. 
% Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters for Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The approach of mode shapes to each other for these small values of $\delta/H$ indicates that the dense vegetation asymptote is reached. 
Mode 1 shapes appear self-similar in shape as $\delta\to 0$.
Inset shows rescaled $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$ approach a universal shape, indicating that an asymptotic limit has been reached. 
% The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the vegetation height in this case is comparable to the boundary layer thickness.
}
\label{Asymptotic_mode}
\end{figure}
Similarly here $\Rey/\Ndg$ is the only remaining parameter in \eqref{eqn:mode2asymp}, the mode shape converges in the aforementioned limit, in agreement with our numerical results shown in Fig.~\ref{Asymptotic_mode}. 
The parameter $\Rey/\Ndg$ also sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$; see Fig.~\ref{Re_vs_delta} for comparison with numerical results).
% We interpret Mode 2 as the instability of an inviscid fluid, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass top plays no role. 
% On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

@@

>>>>

Next we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in ~\eqref{navier-stokes} and expand to linear order to investigate the evolution of small perturbations $(\tilde{u}, \tilde{v})$, which obey
\begin{equation}
\begin{split}
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0,
\end{split} \nonumber
\end{equation}
where the tilde are dropped.
These equations are non-dimensionalized using half channel height $H$, velocity $U_0$, and time $H/U_0$, leading to three non-dimensional parameters, \textit{viz.} $\Rey$, $\Ndg$, and the vegetation submergence ratio $\hg/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the vegetation density and help elucidate the instability mechanism. 
Using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a solution of 
the form $\left(u,v,\psi \right)= \left(\hat u(y), \hat v(y), \phi(y) \right)e^{ikx+\sigma t}$ to obtain a modified Orr-Sommerfield equation \cite{Drazin81} 
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+\Rey \Ndg D\left(2 S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$. 
The growth rate $\sigma$ for a given wave number $k$ appears as an eigenvalue that allows a non-trivial solution $\phi$ of  \eqref{Orr-somerfield}.
We solve \eqref{Orr-somerfield} numerically for $\sigma$ and $\phi$.

@@

\begin{table*}
% \footnotesize
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& Equation \eqref{eqn:Rayleigh}		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  no grass included & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& yes, near $y=\hg$			& no, spans water column
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two unstable modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the instability mechanism, we consider the dependence of the fastest growing wavenumber on $\delta$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical $\delta$ discontinuously jumps and remains $O(1)$ (see Fig.~\ref{Re_vs_delta}). 
To aid in explaining this behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in Fig.~\ref{K_Re_sigma_set3}. 
The smallest $\Rey$ on the neutral curve (Re$(\sigma)=0$) sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two; we refer to the region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

@@

\subsection{Mode 2}
The threshold condition for Mode 2 is numerically observed to be $\Rey \propto ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in Fig.~\ref{Re_vs_delta}, which can be understood by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
% Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the vegetation is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
In this limit, the non-dimensional flow in the grass bed is $U\sim (\Rey \Ndg)^{-1/2} \ll 1$, and therefore $\sigma + ikU \approx \sigma$ to leading order. 
Furthermore, $U_{yy}$ decays to zero outside the boundary layer. 
Outside the grass, the turbulent viscous stress term is negligible compared to the inertial term. Thus, \eqref{Orr-somerfield} simplifies to 
\begin{subequations}
\begin{align}
% \begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for } y<\hg  \label{eqn:mode2asympa} \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for } y>\hg. \label{eqn:mode2asympb}
% \end{split}
\end{align}
\label{eqn:mode2asymp}
\end{subequations}

\begin{figure}
\includegraphics[]{Asymptotic_noshear}
% \includegraphics[width=12cm]{Asymptotic_noshear}
\caption{
Plot of the neutral Mode 1 (solid) and Mode 2 (dashed) shape $|\phi|$ in the limit of small $\delta/H$ for $\hg/H=0.2$. 
% Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters for Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The approach of mode shapes to each other for these small values of $\delta/H$ indicates that the dense vegetation asymptote is reached. 
Mode 1 shapes appear self-similar in shape as $\delta\to 0$.
Inset shows rescaled $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$ approach a universal shape, indicating that an asymptotic limit has been reached. 
% The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the vegetation height in this case is comparable to the boundary layer thickness.
}
\label{Asymptotic_mode}
\end{figure}
The only remaining parameter in \eqref{eqn:mode2asymp} is $\Rey/\Ndg$. 
For fixed $\Rey/\Ndg$, the mode shape converges in the aforementioned limit, in agreement with our numerical results shown in Fig.~\ref{Asymptotic_mode}.
This convergence indicates that we have identified the correct asymptotic limit to investigate Mode 2.

The parameter $\Rey/\Ndg$ therefore sets the threshold, justifying the numerically observed asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$; see Fig.~\ref{Re_vs_delta} for comparison with numerical results).
The structure of this mode in the aforementioned limit is such that $\phi$ is continuous at $y=h_g$, but $D\phi$ undergoes a rapid transition there on the scale of boundary layer thickness $\delta$.
The eigenvalues and the mode shape are otherwise independent of $\delta$, and therefore we conclude that the boundary layer only plays a secondary role of regularizing the discontinuity in tangential velocity arising at $y=\hg$ in this instability mechanism.
% We interpret Mode 2 as the instability of an inviscid fluid, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass top plays no role. 
% On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

@@

Mode 2 is has characteristics distinct from Kelvin-Helmholtz. Outside the grass, the unstable mode shape is governed by the inviscid Rayleigh's equation
\begin{align}
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi. 
\label{eqn:Rayleigh}
\end{align}
An inflection point in $U(y)$ is necessary condition for instability arising from \eqref{eqn:Rayleigh} according to Rayleigh's criteria \cite{Rayleigh1879}. 
However, for our $U(y)$ profiles, $U_{yy}(y) = -1$, and therefore does not change sign $y>\hg$. 
Instead, the dynamics are coupled with the flow in the grass bed described by \eqref{eqn:mode2asympa} in $y< \hg$.
The absence of $U_{yy}$ in \eqref{eqn:mode2asympa} indicates that $U_{yy}$ is approximated to be zero in $y<\hg$, and therefore the negative values of $U_{yy}$ that occurs in the boundary layer do not affect this mode of instability to leading order.
Furthermore, the presence of the critical parameter $\Rey/\Ndg$ in \eqref{eqn:mode2asympa} indicates the presence of alternative destabilizing dynamics, involving the interaction of flow in the unvegetated region governed by \eqref{eqn:Rayleigh} with the flow in the vegetated region incorporating the drag.
Based on this analysis, we conclude that Mode 2 is distinct from the Kelvin-Helmholtz instability, and bears its existence purely because of the vegetation drag.
